---
title: "Practical Machine Learning Course Project"
output: html_document
author: Patrick Meyer
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load data (background), echo=FALSE}
load("D:/NextCloud/Wissenschaft/Practical Machine Learning/Course Project/GlobalEnvironment.RData")
```

```{r Load data}
training = read.csv(file = "C:/Users/Patrick/Desktop/pml-training.csv")
testing = read.csv(file = "C:/Users/Patrick/Desktop/pml-testing.csv")
```


```{r Introduction}

```

```{r Loading libraries, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
library(caret)
```
After loading the data and the necessary packages we will take a close look at the data and preprocess it.
In its raw form the training data has 160 variables with 19622 observations each. However, 67 of these variables consist of more than 97% missing values while the remaining 93 variables do not contain missing values at all. A similar case can be made for the testing data. Therefore, variables with missing values will be excluded.
```{r Data Wrangling 1 - Missing values}
#Check the dimensions of the training data
dim(training)

#Check for variables with and without missing values and exclude those with missing values
sum(apply(training, 2, function(x) sum(is.na(x))/nrow(training) > 0.97))
sum(apply(training, 2, function(x) sum(is.na(x))/nrow(training) == 0))
training = training[,colSums(is.na(training)) == 0]

#Check the dimensions of the testing data
dim(testing)

#Repeat the process for the testing data
sum(apply(testing, 2, function(x) sum(is.na(x))/nrow(testing) == 1))
sum(apply(testing, 2, function(x) sum(is.na(x))/nrow(testing) == 0))
testing = testing[,colSums(is.na(testing)) == 0]
```
Next, we will remove all predictors from both sets with a variance close to zero since we can assume that a predictor with very little variance does not provide sufficient predictive value. This further reduces the number of variables to 59 for each data set.
```{r Data Wrangling 2 - near zero variance predictors}
#Check for cols with near zero variance and exclude them
training = training[,-nearZeroVar(training)]
testing = testing[,-nearZeroVar(testing)]

dim(training)
dim(testing)
```
Since we are interested in predicting the execution of a workout repetition with motion sensors we will further exclude variables that do not fit this description. This leaves us with 53 variables in each set.
```{r Data Wrangling 3 - exclude time variables and id}

head(training[1:6])
head(testing[1:6])

training = training[,-(1:6)]
testing = testing[,-(1:6)]
```
As the next step of the preprocessing the classe variable we are interested in is turned into a factor variable for correct handling by the training methods.
```{r Data Wrangling 4 - turn classe in to a factor variable}
training$classe = as.factor(training$classe)

```
After the previous adjustments the training data can be split into a training set for the actual training and a validation set to test the accuracy of different algorithms.
```{r Splitting the data}
inBuild = createDataPartition(training$classe, p = 0.75, list = FALSE)
validation = training[-inBuild,]
training = training[inBuild,]

```

```{r Creating different models}

```

```{r Predicting the outcome with each model}
#rpartPred = predict(rpartFit, newdata = validation)
```

```{r Accuracy of the rpart model}
#confusionMatrix(testing$classe, rpartPred)
```

```{r Accuracy of the lda model}
#confusionMatrix(testing$classe, ldaPred)
```

```{r Accuracy of the nb model}
#confusionMatrix(testing$classe, nbPred)
```

```{r Accuracy of the gbm model}
#confusionMatrix(testing$classe, gbmPred)
```

```{r Accuracy of the rf model}
#confusionMatrix(testing$classe, rfPred)
```

```{r Accuracy of the rf with cross validation model}
#confusionMatrix(testing$classe, rfPredControl)
```

```{r Applying the best model to the finalTest set}

```




